---
title: "Data visualization"
author: "Cecile Murray"
output: html_document
---

```{r setup, include=FALSE}
libs <- c("here", "tidyverse", "magrittr", "knitr", "kableExtra", "janitor", "tidycensus")
lapply(libs, library, character.only = TRUE)
```

## Lesson goals


- Talk about effective data viz
- Demonstrate a different way to get data (Census API)
- Introduce ggplot2 and "the grammar of graphics"

## Essential parts of a chart

- declarative title
- descriptive subtitle
- axis labels
- source
- often, a legend 

[Make "bad" vs "good charts]

## Getting data from an API

API = "Application-Program Interface"

APIs are how software programs give each other data in mutually intelligible formats. These formats are generally not very human-readable, but data structures are consistent, which is good for programming.

Data analysts like APIs because they remove the manual step of going to a website, clicking download, and moving the file to the right working directory.

### Census API

There is a very nice R package called `tidycensus` that makes getting US Census data very easy. 
This is how easy it is:

```{r census_pull}
# get median household income of each county in the US
medinc <- get_acs("county",
                  table = "B01001",
                  year = 2018)

```

**Bonus content:** 