---
title: "Intro to Data Analysis in R for Public Policy"
author: "Cecile Murray"
date: "12/16/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
libs <- c("here", "tidyverse", "magrittr", "knitr", "kableExtra", "janitor")
lapply(libs, library, character.only = TRUE)
```

## Course goals

- Learn how to perform basic data analysis tasks in R
- Learn some fundamental programming concepts that will help with learning new tools
- Learn *when* and *how* writing code is better than Excel: choosing the right tool for the job

## About me

- 2nd year MSCAPP student at Harris
- Formerly:
  - Senior Research Assistant at The Brookings Institution's Metropolitan Policy Program
  - Research Assistant at the Center on Budget and Policy Priorities (CBPP)
- Alumna of the College: 
  - econ + public policy AB'15 
  - UCIPSS scholar

## Overview of today

1. Talk about when doing data analysis programmatically is advantageous
2. Introduce the tools we'll be using in this series

## Why write code?

Some advantages:

- easier to check (in some ways)
- reproducible
- scalable
- perform more sophisticated analyses (e.g. regression, machine learning)

Some disadvantages:

- less accessible to broad audiences
- can be more opaque
- can be more time-consuming
- more vulnerable to certain kinds of errors

## About learning to code

- Above all, it takes practice (and a lot of Googling)
- Understanding key concepts and terms helps guide that self-directed learning

## Why use R?

```{r whyR, echo=FALSE, warning=FALSE}
use_cases <- c("Free?",
               "Nice user interface",
               "Ease of data assembly",
               "Ease of descriptive stats",
               "Quality of visualization",
               "More complex modeling",
               "General-purpose programming")

tools <- c("Excel", "Stata", "Python", "R")

r <- c("Yes", "Yes", "Excellent", "Very good", "Excellent", "Yes", "Yes")
excel <- c("No*", "Yes", "Good", "Very good", "Good", "No", "No")
stata <- c("No", "Yes", "Good", "Excellent", "Mediocre", "Yes", "No")
python <- c("Yes", "No", "Good", "Good", "Good", "Yes", "Yes")

# make a table out of these
tibble(use_cases, excel, stata, python, r) %>% 
  kable(caption = "Comparing common data analysis tools", 
        col.names = c("", tools)) %>% 
  kable_styling(latex_options = "HOLD") %>% 
  add_footnote("*But your organization probably already paid for MS Office")
```

## R vs. RStudio

R = the programming language

RStudio = an integrated development environment (IDE) for R, aka a nice interface





