---
title: "Summarizing data with dplyr and the tidyverse"
author: "Cecile Murray"
date: "2/7/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
libs <- c("here",
          "tidyverse",
          "magrittr",
          "knitr", 
          "kableExtra")
lapply(libs, library, character.only = TRUE)
set.seed(2020)
```

## Goals for today

- Go over some key terms and syntax
- Introduce some typical data cleaning concepts and tools
- Introduce basic commands to compute descriptive statistics and how to join two datasets
- Use these on real data

## Key terms

- Variable: refers to the columns of a dataframe (usually).
- Observation: refers to one unit, often one row, of a dataframe

```{r df_terms, echo=FALSE}
key_term_df <- tibble(
  "person_id" = str_pad(as.character(seq(1, 4)), width = 2),
  "age" = round(runif(4, 25, 85)),
  "sex" = c("F", "F", "M", "M")
)

kable(key_term_df, align="l")
```

## tidyverse and dplyr verbs

- `dplyr` is a package in the `tidyverse` family of packages
- Common "grammar" for working with data, objects, functions, etc

## Data cleaning: common tasks

- converting variable types: string to numeric and vice versa, dates, etc
- look for missing values
- "recode" variables into useful categories
- reshape the data: convert columns and rows
- look up values in another dataset

## **mutate**

```{r mutate, echo=FALSE}
key_term_df %>% kable(align = "l")

key_term_df %<>% 
  mutate("educ" = c("HS", "HS", "BA", "BA")) 

key_term_df %>% kable(align = "l")

```

## **filter**

```{r filter, echo=FALSE}
key_term_df %>% kable(align = "l")

key_term_df %>% 
  filter(sex == "F") %>% 
  kable(align = "l")

```

## **select**

```{r select, echo=FALSE}
key_term_df %>% kable(align = "l")

key_term_df %>% 
  select(person_id, age) %>% 
  kable(align = "l")

```

## Summary statistics: examples

- number of reported crimes of each type
- % resulting in an arrest for each type of crime
- average number of crimes by day of week

## Merging tables with joins

- **inner join**: keep only rows that have a match in both tables
- **left join**: keep all rows in left table, discard rows in right table with no match
- **right join**: keep all rows in right table, discard rows in left table with no match
- **outer join**: keep all rows, regardless of whether there is a match

## left_join or right_join

```{r left_join, echo=FALSE}
occp_df <- key_term_df %>% 
  select(person_id) %>% 
  mutate("occupation" = c("doctor", "lawyer", "engineer", "NA")) 
occp_df %>% kable(align = "l") 

occp_df %>% 
  left_join(key_term_df, by = "person_id") %>% 
  kable(align = "l")
```

## inner_join

```{r inner_join, echo=FALSE}
occp_df <- key_term_df %>% 
  select(person_id) %>% 
  mutate("occupation" = c("doctor", "lawyer", "engineer", "NA"))

occp_df %>% 
  left_join(key_term_df, by = "person_id") %>% 
  kable(align = "l")
```

## full_join (aka outer join)

```{r full_join, echo=FALSE}
occp_df <- key_term_df %>% 
  select(person_id) %>% 
  mutate("occupation" = c("doctor", "lawyer", "engineer", "NA"))

occp_df %>% 
  left_join(key_term_df, by = "person_id") %>% 
  kable(align = "l")
```

